Analysis and elimination of bottlenecks in parallel algorithm for solving global optimization problems

### 2022-04-15 2022/RSD/paper.tex
minor corrections

The main part of the paper has the following structure. In Section 2, the global optimization problem statement is presented and the properties of the objective function allowing developing the efficient global minimum search algorithms are described. In Section 3, the scheme of the parallel global optimization algorithm is presented and the theoretical estimates of its speedup are given. In Section 4, two local optimization methods (BFGS method and Hooke-Jeeves one) are considered, which can be applied for refining the solution found. Possible parallelization schemes for these methods are discussed. Section 5 contains the results of the computational experiments conducted using both the test problems and the applied ones. Section 6 concludes the paper.

\section{Global optimization problem statement}

@@

In the theory, the Peano-Hilbert curve is a limit object. Therefore, one needs to construct an approximation to the theoretical curve for the practical application. The issues of numerical construction of such approximation (also called \textit{evolvents}) for given dimensionality with given precision were considered in \cite{Sergeyev2013}.

The use of the space-filling curves allows reducing the multidimensional problem \ref{problemN} to a one-dimensional problem and applying the efficient one-dimensional optimization methods to solve it, i.e. 
\begin{equation} \label{problem_ref}
\varphi(y^*) = \varphi(y(x^*)) = \min{\{ \varphi(y(x)): x \in [0, 1] \}}.
\end{equation}
Keeping the property of limited relative differences of the objective function is an important property of using the evolvents. If the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, the function $\varphi(y(x))$ in the interval $[0,1]$ will satisfy the uniform H{\"o}lder condition
\begin{equation} \label{holder_ref}
\left| \varphi(y(x_1)) - \varphi(y(x_2)) \right| \leq H {\left| x_1 - x_2 \right|}^{1/N}, \; x_1, x_2 \in [0,1],
\end{equation}
where H{\"o}lder constant $H$ is related to Lipschitz constant $L$ by the relation $H = 2L\sqrt{N+3}$. Therefore, without liming the generality, one can consider the minimization of the one-dimensional function 
\begin{equation} \label{minim_fun_ref}
f(x) = \varphi(y(x)), x \in [0,1],
\end{equation}
satisfying H{\"o}lder condition.

@@

Step 5. Carry out new trials in parallel at the points $x^{k+j}$, $1 \leq j \leq p$, computed by the formulae
$$ x^{k+j} = \frac{x_{t_j} + x_{t_j - 1}}{2} - sign(z_{t_j} - z_{t_j - 1}) \frac{1}{2r} {\Bigg[\frac{\left| z_{t_j} - z_{t_j - 1} \right|}{\mu}\Bigg]}^{N}.$$
Put the results of the trials performed into the information database of the algorithm.

Step 6. Check the stopping criterion $\Delta_{t_j} \leq \varepsilon$, where $t_j$ are from Step 4 and $\varepsilon > 0$ is the predefined accuracy. 
After completing the algorithm work, take as the estimate of the solution of the problem \ref{problemN} 
$$ f_k^* = \min\limits_{1 \leq i \leq k} f(x^i), \; x_k^* = \arg \min\limits_{1 \leq i \leq k} f(x^i).$$

The theoretical substantiation of the presented algorithm (denoted hereafter as \textit{global search algorithm} or GSA) see in more details in \cite{Strongin2000}. Note briefly that the characteristics of the intervals (\ref{R}) used in the algorithm can be considered as a quantitative measure how particular interval is promising from the viewpoint of finding the global minimum inside it. The inequalities (\ref{OrderedR}) arrange the intervals according to their characteristics, and the trials are performed in parallel in $p$ most promising intervals.

@@

Let us describe (according to \cite{Strongin2000}) the theoretical properties of the parallel algorithm featuring its speedup. In the problems being solved, the time of executing a single search trial essentially exceeds the time of processing its result. Therefore, the speedup in the number of trials 
\begin{equation} \label{par_trl_ref}
s(p) = \frac{n(1)p}{n(p)}
\end{equation}
is the key characteristic of the efficiency of the parallel global search algorithm.
Here $n(1)$ is the number of trials executed by the sequential method and $n(p)$ is the number of trials executed by the parallel method employing $p$ parallel processes.

@@

\textbf{Consequence}. Let the objective function $f(x)$ have $Q$ points of local minima $\{x_1^{\prime},...,x_Q^{\prime}\}$, which the condition \eqref{first_s_ref} is fulfilled for, and let there exist the trial points $y^{n_i}$, $1 \leq i \leq Q$ such as 
\begin{gather} 
    y^{n_i} \in \{y^1,...y^{q(l)}\}, \nonumber \\ 
    \alpha_i \leq y^{n_i} \leq \alpha_{i+1}, \; \alpha_i, \alpha_{i+1} \in \{x^*, x_1^{\prime},...,x_Q^{\prime}\}, 1 \leq i \leq Q. \nonumber
\end{gather}
Then, if the conditions of theorem are fulfilled, the parallel algorithm with the parallelization degree $Q+1$ will be non-redundant (i.e. $s(Q+1)=Q+1, \; \lambda(Q+1)=0$) until the condition \eqref{p_two_ref} is fulfilled.

The consequence from theorem plays a special role in solving the multidimensional problems reduced to the one-dimensional ones using the evolvent $y(x)$. The evolvent $y(x)$, being an approximation to the Peano curve, has the effect of ``splitting'' of the global minimum point $y^* \in D$ into several (up to $2^N$) preimages in the interval [0,1]. Applying the parallel algorithm to minimize the reduced one-dimensional function, one can expect a zero redundancy at the parallelization degree up to $2^N+1$.

Now, let us consider the case when after the stage of the search for the rough approximation to the global solution, the stage of its local refinement is performed. Assume that the fraction of the search trials necessary for the local refinement of the solution makes $\alpha$ of total number of trials. Then, according to Amdahl's law, the overall speedup $S(p)$, which can be obtained by parallelization into $p$ processes, cannot exceed the quantity 
$$ S(p) = {\Big(\alpha + \frac{1-\alpha}{p}\Big)}^{-1}.$$

@@

The general scheme of the parallel computing for all the local methods employed is presented in Fig. \#. The objective function values can be computed on a single node using shared memory (OpenMP), on a cluster with distributed memory (MPI), or on accelerator using CUDA.

\section{Results of numerical experiments}

@@

The first series of experiments was carried out with the local optimization algorithms only. Rosenbrock function is a classical test function for the local search methods 
\begin{gather}
    \varphi(y) = \sum_{i=1}^{N-1} \left((1-y_i)^2 + 100(y_{i+1} - y_i^2)^2\right) \nonumber \\
    D = \{ y \in R^N: -25 \leq y_j \leq 25, 1 \leq j \leq N \}. \nonumber
\end{gather}

@@

\begin{table}[ht]
	\caption{Averaged number of iterations for GSA algorithm}
	\label{tab:5}
	\center
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline
		N & Problem class & P = 1 & P = 2 & P = 4 & P = 8 & P = 16 \\
		\hline 
		    \multirow{2}{*}{4} & hard & 36528 & 18651 & 9854 & 4808 & 2260 \\ \cline{2-7}
		                       & simple & 9995 & 5199 & 2706 & 1333 & 630 \\
		\hline
		    \multirow{2}{*}{5} & hard & 266420 & 149109 & 90894 & 39873 & 22206 \\ \cline{2-7}
		                       & simple & 29853 & 12913 & 7296 & 3618 & 2220 \\
		\hline
	\end{tabular}
\end{table}

>>>>

The main part of the paper has the following structure. In Section \ref{sec:problem}, the global optimization problem statement is presented and the properties of the objective function allowing developing the efficient global minimum search algorithms are described. In Section \ref{sec:global}, the scheme of the parallel global optimization algorithm is presented and the theoretical estimates of its speedup are given. In Section \ref{sec:local}, two local optimization methods (BFGS method and Hooke-Jeeves one) are considered, which can be applied for refining the solution found. Possible parallelization schemes for these methods are discussed. Section \ref{sec:results} contains the results of the computational experiments conducted using both the test problems and the applied ones.
% Section 6 concludes the paper.

\section{Global optimization problem statement}\label{sec:problem}

@@

In theory, the Peano-Hilbert curve is a limit object. Therefore, one needs to construct an approximation to the theoretical curve for the practical application. The issues of numerical construction of such approximation (also called \textit{evolvents}) for given dimensionality with given precision were considered in \cite{Sergeyev2013}.

The use of the space-filling curves allows reducing the multidimensional problem (\ref{problemN}) to a one-dimensional problem and applying the efficient one-dimensional optimization methods to solve it, i.e. 
\begin{equation} \label{problem_ref}
\varphi(y^*) = \varphi(y(x^*)) = \min{\{ \varphi(y(x)): x \in [0, 1] \}}.
\end{equation}
Keeping the property of limited relative differences of the objective function is an important feature of the evolvents. If the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, the function $\varphi(y(x))$ in the interval $[0,1]$ will satisfy the H{\"o}lder condition
\begin{equation} \label{holder_ref}
\left| \varphi(y(x_1)) - \varphi(y(x_2)) \right| \leq H {\left| x_1 - x_2 \right|}^{1/N}, \; x_1, x_2 \in [0,1],
\end{equation}
where H{\"o}lder constant $H$ is related to Lipschitz constant $L$ by the relation $H = 2L\sqrt{N+3}$. Therefore, without liming the generality, one can consider the minimization of the one-dimensional function 
\begin{equation} \label{minim_fun_ref}
f(x) = \varphi(y(x)), \; x \in [0,1],
\end{equation}
satisfying H{\"o}lder condition.

@@

Step 5. Carry out new trials in parallel at the points $x^{k+j}$, $1 \leq j \leq p$, computed by the formula
$$ x^{k+j} = \frac{x_{t_j} + x_{t_j - 1}}{2} - sign(z_{t_j} - z_{t_j - 1}) \frac{1}{2r} {\Bigg[\frac{\left| z_{t_j} - z_{t_j - 1} \right|}{\mu}\Bigg]}^{N}.$$
Put the results of the trials performed into the information database of the algorithm.

Step 6. Check the stopping criterion $\Delta_{t_j} \leq \varepsilon$, where $t_j$ are from Step 4 and $\varepsilon > 0$ is the predefined accuracy. 
After completing the algorithm work, take as the estimate of the solution of the problem (\ref{problemN})
$$ f_k^* = \min\limits_{1 \leq i \leq k} f(x^i), \; x_k^* = \arg \min\limits_{1 \leq i \leq k} f(x^i).$$

The theoretical basis of the presented algorithm (denoted hereafter as \textit{global search algorithm} or GSA) see in more details in \cite{Strongin2000}. Note briefly that the characteristics of the intervals (\ref{R}) used in the algorithm can be considered as a quantitative measure how particular interval is promising from the viewpoint of finding the global minimum inside it. The inequalities (\ref{OrderedR}) arrange the intervals according to their characteristics, and the trials are performed in parallel in $p$ most promising intervals.

@@

Let us describe (according to \cite{Strongin2000}) the theoretical properties of the parallel algorithm featuring its speedup. In the problems being solved, the time of executing a single search trial essentially exceeds the time of processing its result. Therefore, the speedup in the number of trials 
\begin{equation} \label{par_trl_ref}
s(p) = \frac{n(1) \cdot p}{n(p)}
\end{equation}
is the key characteristic of the efficiency of the parallel global search algorithm.
Here $n(1)$ is the number of trials executed by the sequential method and $n(p)$ is the number of trials executed by the parallel method employing $p$ parallel processes.

@@

\textbf{Consequence}. Let the objective function $f(x)$ have $Q$ points of local minima $\{x_1^{\prime},...,x_Q^{\prime}\}$, which the condition \eqref{first_s_ref} is fulfilled for, and let there exist the trial points $y^{n_i}$, $1 \leq i \leq Q,$ such that 
\begin{gather} 
    y^{n_i} \in \{y^1,...y^{q(l)}\}, \nonumber \\ 
    \alpha_i \leq y^{n_i} \leq \alpha_{i+1}, \; \alpha_i, \alpha_{i+1} \in \{x^*, x_1^{\prime},...,x_Q^{\prime}\}, 1 \leq i \leq Q. \nonumber
\end{gather}
Then, if the conditions of theorem are fulfilled, the parallel algorithm with the parallelization degree $Q+1$ will be non-redundant (i.e. $s(Q+1)=Q+1, \; \lambda(Q+1)=0$) until the condition \eqref{p_two_ref} is fulfilled.

The consequence from the theorem plays a special role in solving the multidimensional problems reduced to the one-dimensional ones using the evolvent $y(x)$. The evolvent $y(x)$, being an approximation to the Peano curve, has the effect of ``splitting'' of the global minimum point $y^* \in D$ into several (up to $2^N$) preimages in the interval [0,1]. Applying the parallel algorithm to minimize the reduced one-dimensional function, one can expect a zero redundancy at the parallelization degree up to $2^N+1$.

Now, let us consider the case when, after the stage of the search for the rough approximation to the global solution, the stage of its local refinement is performed. Assume that the fraction of the search trials necessary for the local refinement of the solution makes $\alpha$ of total number of trials. Then, according to Amdahl's law, the overall speedup $S(p)$, which can be obtained by parallelization into $p$ processes, cannot exceed the quantity 
$$ S(p) = {\Big(\alpha + \frac{1-\alpha}{p}\Big)}^{-1}.$$

@@

%The general scheme of the parallel computing for all the local methods employed is presented in Fig. \#. The objective function values can be computed on a single node using shared memory (OpenMP), on a cluster with distributed memory (MPI), or on accelerator using CUDA.
Thus, in both considered local methods the parallelization scheme will be generally the same. The vector of objective function values can be computed in parallel on a single node using shared memory (OpenMP), on a cluster with distributed memory (MPI), or on accelerator using CUDA.

\section{Results of numerical experiments}\label{sec:results}

@@

The first series of experiments was carried out with the local optimization algorithms only. Rosenbrock function is a classical test function for the local search methods 
\begin{gather}
    \varphi(y) = \sum_{i=1}^{N-1} \left[(1-y_i)^2 + 100(y_{i+1} - y_i^2)^2\right], \nonumber \\
    D = \{ y \in R^N: -25 \leq y_j \leq 25, 1 \leq j \leq N \}. \nonumber
\end{gather}

@@

\begin{table}[ht]
	\caption{Averaged number of iterations for GSA algorithm}
	\label{tab:5}
	\center
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline
		$N$ & Problem class & $p=1$ & $p=2$ & $p=4$ & $p=8$ & $p=16$ \\
		\hline 
		    \multirow{2}{*}{4} & Hard & 36528 & 18651 & 9854 & 4808 & 2260 \\ \cline{2-7}
		                       & Simple & 9995 & 5199 & 2706 & 1333 & 630 \\
		\hline
		    \multirow{2}{*}{5} & Hard & 266420 & 149109 & 90894 & 39873 & 22206 \\ \cline{2-7}
		                       & Simple & 29853 & 12913 & 7296 & 3618 & 2220 \\
		\hline
	\end{tabular}
\end{table}
